---
title: "Exploration of Stream Data"
author: "Julia Piaskowski"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format: 
  html: 
    embed-resources: true
---


## Prep: Libraries, Data

```{r include=FALSE, eval=TRUE}
# standard libraries
library(dplyr); library(tidyr); library(readr); library(ggplot2)

# date & time / time series libraries
library(lubridate)
```

#### Data

Import data set and create some helper variables:  

```{r}
streams <- read_csv(here::here("data/Alpowa_Deadman_Flow_Temp_2022.csv"), 
                    show_col_types = FALSE) |> 
  janitor::clean_names() |> 
  rename(date_ = "date") |> # rename header to 'date_' since 'date()' is a function
  mutate(missing = case_when(     # create new variable to indicate extent of missingness
    is.na(water_temp_c) ~ "missing",
    .default = NA_character_
  )) |> 
  mutate(year = year(date_), julian = as.numeric(format(date_, "%j"))) |> # new variables: Year, Julian days
  mutate(Year = as.character(year)) |> 
  mutate(improvement = case_when(    # add new variable for magic cut-off
    date_ < "2011-07-01" ~ "before", 
    date_ >= "2011-07-01" ~ "after"
  )) |> 
  dplyr::mutate(improvement = factor(improvement, levels = c("before", "after"))) |> 
  #filter(date_ > "2003-06-30" & date_ < "2022-07-01") |> # filter to interval of interest
  slice(-(grep("02-29$", date_))) 
```

Filter to individual streams:
```{r}
deadman <- filter(streams, site == "Deadman Creek") |> 
  dplyr::arrange(date_) |> 
  mutate(index = 1:n())

alpowa <- filter(streams, site == "Alpowa Creek") |> 
  dplyr::arrange(date_) |> 
  mutate(index = 1:n())
```

Pivot to long format for plotting: 
```{r}
deadman_long <- deadman |> select(date_, julian, flow_cfs, water_temp_c) |> 
  pivot_longer(cols = 3:4)

alpowa_long <- alpowa |> select(date_, julian, flow_cfs, water_temp_c) |> 
  pivot_longer(cols = 3:4)
```


## EDA (exploratory data analysis)

### Alpowa Creek
*(exploratory plots galore)*

```{r fig.asp=1}
hist(log(alpowa$flow_cfs))
hist(alpowa$water_temp_c)
plot(alpowa$flow_cfs, alpowa$water_temp_c)
plot(log(alpowa$flow_cfs), alpowa$water_temp_c)
```

#### General time series
```{r}
ggplot(alpowa_long, aes(x = date_, y = value)) +
  geom_line(color = "dodgerblue3") +
  facet_grid(name ~ ., scales = "free_y") + 
  theme_bw(base_size = 12) +
  ggtitle("Alpowa Creek") + 
  theme(axis.title.x = element_blank())

#ggsave("outputs/Alpowa_temp.png", height = 975, width = 2625, units = "px")
```

#### Time series faceted by year
```{r, fig.height=8, out.height=8}
ggplot(alpowa, aes(x = julian, y = water_temp_c)) +
  geom_line(aes(color = improvement), linewidth = 0.75) +
  facet_grid(rows = vars(Year)) + 
  scale_color_manual(values = c("dodgerblue2", "maroon")) +
  guides(color = "none") + 
  ylab("Water Temperature (째C)") +
  xlab("Julian Calendar Days") + 
  ggtitle("Alpowa Creek") + 
  scale_y_continuous(breaks=c(0, 10, 20)) +
  theme_minimal(base_size = 11) 

#ggsave("outputs/Alpowa_temp_stacked.png", height = 2850, width = 2025, units = "px")
```

```{r, fig.height=8, out.height=8}
ggplot(alpowa, aes(x = julian, y = flow_cfs)) +
  geom_line(aes(color = improvement), linewidth = 0.75) +
  facet_grid(rows = vars(Year), scales = "free_y") + 
  scale_color_manual(values = c("dodgerblue2", "maroon")) +
  guides(color = "none") + 
  ylab("Flow rate (CFS)") +
  xlab("Julian Calendar Days") + 
  ggtitle("Alpowa Creek") + 
  scale_y_continuous(breaks=c(0, 10, 20)) +
  theme_minimal(base_size = 11) 

#ggsave("outputs/Alpowa_temp_stacked.png", height = 2850, width = 2025, units = "px")
```

#### Time series overlaid
```{r}
ggplot(alpowa, aes(x = julian, y = water_temp_c, group = Year)) +
  geom_line(aes(color = improvement), linewidth = 0.5, alpha = 0.6) +
  guides(color = "none") + 
  ylab("Water Temperature (째C)") +
  xlab("Julian Calendar Days") + 
  scale_color_manual(values = c("dodgerblue2", "maroon")) +
  ggtitle("Alpowa Creek") + 
  scale_y_continuous(breaks=c(0, 10, 20)) +
  theme_linedraw(base_size = 11) 
```

```{r}
ggplot(alpowa, aes(x = julian, y = flow_cfs, group = Year)) +
  geom_line(aes(color = improvement), linewidth = 0.5, alpha = 0.6) +
  guides(color = "none") + 
  ylab("Water Flow (CFS)") +
  xlab("Julian Calendar Days") + 
  scale_y_continuous(trans='log10') +
  scale_color_manual(values = c("dodgerblue2", "maroon")) +
  ggtitle("Alpowa Creek", "log10 scale") + 
  theme_linedraw(base_size = 11) 
```


### Deadman Creek
*(exploratory plots galore)*

```{r fig.asp=1}
hist(log(deadman$flow_cfs))
hist(deadman$water_temp_c)
plot(deadman$flow_cfs, deadman$water_temp_c)
plot(log(deadman$flow_cfs), deadman$water_temp_c)
```

#### General time series

```{r}
ggplot(deadman_long, aes(x = date_, y = value)) +
  geom_line(color = "dodgerblue3") +
  facet_grid(name ~ ., scales = "free_y") + 
  theme_bw(base_size = 12) +
  ggtitle("Deadman Creek") + 
  theme(axis.title.x = element_blank())

#ggsave("outputs/Alpowa_temp.png", height = 975, width = 2625, units = "px")
```

#### Time series faceted by year

```{r, fig.height=8, out.height=8}
ggplot(deadman, aes(x = julian, y = water_temp_c)) +
  geom_line(aes(color = improvement), linewidth = 0.75) +
  facet_grid(rows = vars(Year)) + 
  scale_color_manual(values = c("dodgerblue2", "maroon")) +
  guides(color = "none") + 
  ylab("Water Temperature (째C)") +
  xlab("Julian Calendar Days") + 
  ggtitle("Deadman Creek", "Water Temp") + 
  scale_y_continuous(breaks=c(0, 10, 20)) +
  theme_minimal(base_size = 11) 

#ggsave("outputs/Alpowa_temp_stacked.png", height = 2850, width = 2025, units = "px")
```

```{r, fig.height=8, out.height=8}
ggplot(deadman, aes(x = julian, y = flow_cfs)) +
  geom_line(aes(color = improvement), linewidth = 0.75) +
  facet_grid(rows = vars(Year), scales = "free_y") + 
  scale_color_manual(values = c("dodgerblue2", "maroon")) +
  guides(color = "none") + 
  ylab("Flow rate (CFS)") +
  xlab("Julian Calendar Days") + 
  ggtitle("Deadman Creek", "Flow Rate") + 
  scale_y_continuous(breaks=c(0, 10, 20)) +
  theme_minimal(base_size = 11) 

#ggsave("outputs/Alpowa_temp_stacked.png", height = 2850, width = 2025, units = "px")
```

#### Time series overlaid

```{r}
ggplot(deadman, aes(x = julian, y = water_temp_c, group = Year)) +
  geom_line(aes(color = improvement), linewidth = 0.5, alpha = 0.6, na.rm=TRUE) +
  guides(color = "none") + 
  ylab("Water Temperature (째C)") +
  xlab("Julian Calendar Days") + 
  scale_color_manual(values = c("dodgerblue2", "maroon")) +
  ggtitle("Alpowa Creek") + 
  scale_y_continuous(breaks=c(0, 10, 20)) +
  theme_linedraw(base_size = 11) 
```

```{r}
ggplot(deadman, aes(x = julian, y = flow_cfs, group = Year)) +
  geom_line(aes(color = improvement), linewidth = 0.5, alpha = 0.6, na.rm=TRUE) +
  guides(color = "none") + 
  ylab("Water Flow (CFS)") +
  xlab("Julian Calendar Days") + 
  scale_y_continuous(trans='log10') +
  scale_color_manual(values = c("dodgerblue2", "maroon")) +
  ggtitle("Deadman Creek", "log10 scale") + 
  theme_linedraw(base_size = 11) 
```
