---
title: "Miscellaneous Appproaches"
author: "Julia Piaskowski"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format: 
  html: 
    embed-resources: true
---


## Preparation

Load libraries

```{r include=FALSE, eval=TRUE}
# standard libraries
library(dplyr); library(tidyr); library(readr); library(ggplot2)

# date & time / time series libraries
library(lubridate)
library(forecast)
library(imputeTS)
#library(fable) # ETS()
```


Import prepared data sets:  

```{r}
load(here::here("outputs", "prepped_stream_data.RData"))
```

## Other Ideas

1. what happened if we forecast beyond 2011? (pretend time series ends there) - this doesn't make sense
1. Lag work - why?
1. look at July 1 - Sept 30 (?) as a 'low flow' period and see if the flow rate has changed (Julian days 182 - 273)

### Alpowa flow

```{r}
alpowa_low <- alpowa |> filter(julian > 181 & julian < 274)

alpowa_ts_temp <- ts(alpowa_low$water_temp_c, start = 1, frequency = 92) 
alpowa_ts_flow <- ts(alpowa_low$flow_cfs, start = 1, frequency = 92) 

alpowa_ts_temp_imp <- na_seadec(alpowa_ts_temp, algorithm = "interpolation")
alpowa_ts_flow_imp <- na_seadec(alpowa_ts_flow, algorithm = "interpolation")
```

```{r}
alpowa_flow_stl <- stl(alpowa_ts_flow_imp, 
                      s.window = "periodic", 
                     t.window = 100,  
                      robust = TRUE)  

autoplot(alpowa_flow_stl)  + #year 8 is when the cut-off begins
  geom_vline(xintercept = 8, color = "red")+
  labs(title = "Alpowa Flow")
```

### Alpowa temperature

```{r}
alpowa_temp_stl <- stl(alpowa_ts_temp_imp, 
                      s.window = "periodic", 
                     t.window = 100,  
                      robust = TRUE)  

autoplot(alpowa_temp_stl)  + #year 8 is when the cut-off begins
  geom_vline(xintercept = 8, color = "red")+
  labs(title = "Alpowa Flow")
```

### Mann-Kendall test

```{r}
alpowa_low$alpowa_temp_stl_loess <- alpowa_temp_stl$time.series[,2]
alpowa_low$alpowa_flow_stl_loess <- alpowa_flow_stl$time.series[,2]

cor.test(alpowa_low$alpowa_temp_stl_loess, alpowa_low$index, method = "kendall")
cor.test(alpowa_low$alpowa_flow_stl_loess, alpowa_low$index, method = "kendall")
```


### Both Variables

```{r}
plot(alpowa_low$flow_cfs, alpowa_low$water_temp_c)
cor(alpowa_low$flow_cfs, alpowa_low$water_temp_c, use = "pairwise.complete.obs")
summary(lm(water_temp_c ~ flow_cfs, data = alpowa_low))
```

1. multiple variable forecasting (air temp + water flow on water temp). 










